Parsing Postfix log files

### 2007-06-18 doc/logparser.tex
Bunch of warnings from chktex fixed.

XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS.  REVIEW
EACH PAPER, WRITING AN ABSTRACT FOR IT: WHAT'S THE SAME, WHAT'S DIFFERENT,
PERFORMANCE AND \newline COMPLETENESS COMPARISON.  FROM THIS I'LL GET AN
OVERALL PICTURE, AND I CAN WORK REFERENCES TO THE PAPERS INTO THIS PAPER,
PLUS MAYBE A PRIOR ART SECTION.

XXX TURN OFF RESTRICTIONS FOR ME AND SEE HOW WELL SPAMASSASSIN DEALS
WITH THE RESULTING MAILS; FALSE POSITIVES, FALSE NEGATIVES, ETC.

@@

    \item [server\_hostname] The hostname of the server, it will be
        \textit{unknown} if the IP address could not be resolved to a
        hostname via DNS.

@@

    \item [client\_hostname] The hostname of the client, it will be
        \textit{unknown} if the IP address could not be resolved to a
        hostname via DNS.

@@

    \item [queueid] The queueid of the mail, or \textit{NOQUEUE} if a mail
        wasn't accepted.

@@

Every log line corresponding to Postfix performing an action other than
INFO will have an entry in the results table, e.g. rejecting an SMTP
command, delivering a mail, or bouncing a mail.  Any log line where the
Postfix action is INFO is not interesting in and of itself, but provides
additional information which will be saved with other results.

@@

    \item [warning] Postfix can log a warning instead of rejecting an SMTP
        command - a facility that is quite useful for testing new
        restrictions.  This field will be 1 if the log line parsed was a
        warning rather than a real rejection, or 0 for a real rejection or
        any other result.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g. the rejection message from an RBL.

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g. \texttt{check\_client\_access~\nolinebreak
cidr:/etc/postfix/client\_access}\footnote{\texttt{cidr} is the type of the
lookup table.} will check whether the IP address of the connected client
matches the left hand side of each line in the file and will return the
right hand side of the first matching line.  Other restrictions determine
their result in different ways, e.g. \texttt{reject\_rbl\_client
some.rbl.com} checks whether the IP address of the client is present in the
listed RBL.

Each restriction is evaluated to produce a result of \textit{reject\/},
\textit{permit\/}, \textit{dunno\/} or the name of another restriction to
be evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be handled differently from more general cases.  When the result
is another restriction Postfix will evaluate the new restriction, allowing
restrictions to be chosen based on the client IP address, HELO hostname,
sending address, recipient address, etc.  The administrator can define new
restrictions as a list of existing restrictions, allowing arbitrarily long
and complex sequences of lookups and restrictions.

@@

An important property of a parser is how execution time scales with input
size: does it scale linearly, polynomially, or exponentially?
Graph~\ref{execution time vs file size vs number lines} shows the execution
time in seconds, file size in MB and tens of thousands of lines per log
file.  All three lines run roughly parallel, giving a visual impression
that the algorithm scales linearly with input size.  This is borne(XXX) out
by graph~\ref{execution time vs file size vs number lines factor}, which
plots the ratio of file size in MB vs execution time in seconds and number
of lines

@@

    \item [Log Mail Analyser] \cite{log-mail-analyser} attempts to
        correlate Postfix log lines by queueid and produce both a CSV file
        and either a Berkeley DB or MySQL database.  It aims to save
        \textit{date and hour, DNS name and IP address host, mail server IP
        address, sender, receiver and e-mail status (sent, rejected)\/}, so
        its data extraction is reasonably complete.  The parsing is
        hard-coded into the program; extending it requires familiarity with
        a substantial portion of the code.  Producing a CSV file appears to
        be quite limiting, but the intention is to provide a format
        suitable for use with grep(1), awk(1) and the myriad of other Unix
        utilities already available.  Supporting Berkeley DB is also
        limiting, requiring an understanding of the (partially documented)
        database structure and writing a program to extract the required
        data - it appears to be the least useful of the three output
        formats, though the authors of LMA did provide some queries:
        IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT.  Unfortunately
        when the author tested LMA it didn't work properly and after
        exploring the code for some time the author reluctantly gave up.

>>>>

XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS\@.
REVIEW EACH PAPER, WRITING AN ABSTRACT FOR IT\@: WHAT'S THE SAME, WHAT'S
DIFFERENT, PERFORMANCE AND \newline COMPLETENESS COMPARISON\@.  FROM THIS
I'LL GET AN OVERALL PICTURE, AND I CAN WORK REFERENCES TO THE PAPERS INTO
THIS PAPER, PLUS MAYBE A PRIOR ART SECTION\@.

XXX TURN OFF RESTRICTIONS FOR ME AND SEE HOW WELL SPAMASSASSIN DEALS WITH
THE RESULTING MAILS\@; FALSE POSITIVES, FALSE NEGATIVES, ETC\@.

@@

    \item [server\_hostname] The hostname of the server, it will be
        \textit{unknown\/} if the IP address could not be resolved to a
        hostname via DNS\@.

@@

    \item [client\_hostname] The hostname of the client, it will be
        \textit{unknown\/} if the IP address could not be resolved to a
        hostname via DNS\@.

@@

    \item [queueid] The queueid of the mail, or \textit{NOQUEUE\/} if a
        mail wasn't accepted.

@@

Every log line corresponding to Postfix performing an action other than
INFO will have an entry in the results table, e.g.\ rejecting an SMTP
command, delivering a mail, or bouncing a mail.  Any log line where the
Postfix action is INFO is not interesting in and of itself, but provides
additional information which will be saved with other results.

@@

    \item [warning] Postfix can log a warning instead of rejecting an SMTP
        command --- a facility that is quite useful for testing new
        restrictions.  This field will be 1 if the log line parsed was a
        warning rather than a real rejection, or 0 for a real rejection or
        any other result.

@@

    \item [data] A field available for anything not covered by other
        fields, e.g.\ the rejection message from an RBL\@.

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an SMTP command, one list for each
stage of the SMTP conversation: client connection, HELO command, MAIL FROM
command, RCPT TO command (possibly multiple times), DATA command, and end
of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
uses simple lookup tables as the deciding factor for some restrictions,
e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr} is the type of the lookup table.} will check
whether the IP address of the connected client matches the left hand side
of each line in the file and will return the right hand side of the first
matching line.  Other restrictions determine their result in different
ways, e.g. \texttt{reject\_rbl\_client some.rbl.com} checks whether the IP
address of the client is present in the listed RBL\@.

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno\/} or the name of another restriction to be
evaluated.\footnote{Other results are possible, for full details
see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
\textit{dunno\/} means to stop evaluating the current restriction and
continue with the next restriction in the list, allowing more specific
cases to be handled differently from more general cases.  When the result
is another restriction Postfix will evaluate the new restriction, allowing
restrictions to be chosen based on the client IP address, HELO hostname,
sending address, recipient address, etc.  The administrator can define new
restrictions as a list of existing restrictions, allowing arbitrarily long
and complex sequences of lookups and restrictions.

@@

An important property of a parser is how execution time scales with input
size: does it scale linearly, polynomially, or exponentially?
Graph~\ref{execution time vs file size vs number lines} shows the execution
time in seconds, file size in MB and tens of thousands of lines per log
file.  All three lines run roughly parallel, giving a visual impression
that the algorithm scales linearly with input size.  This is borne out by
graph~\ref{execution time vs file size vs number lines factor}, which plots
the ratio of file size in MB vs execution time in seconds and number of
lines

@@

    \item [Log Mail Analyser]~\cite{log-mail-analyser} attempts to
        correlate Postfix log lines by queueid and produce both a CSV file
        and either a Berkeley DB or MySQL database.  It aims to save
        \textit{date and hour, DNS name and IP address host, mail server IP
        address, sender, receiver and e-mail status (sent, rejected)}, so
        its data extraction is reasonably complete.  The parsing is
        hard-coded into the program; extending it requires familiarity with
        a substantial portion of the code.  Producing a CSV file appears to
        be quite limiting, but the intention is to provide a format
        suitable for use with grep(1), awk(1) and the myriad of other Unix
        utilities already available.  Supporting Berkeley DB is also
        limiting, requiring an understanding of the (partially documented)
        database structure and writing a program to extract the required
        data --- it appears to be the least useful of the three output
        formats, though the authors of LMA did provide some queries:
        IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.
        Unfortunately when the author tested LMA it didn't work properly
        and after exploring the code for some time the author reluctantly
        gave up.
